---
title: "Detecting non alpha character string column and removing leading zeros"
author: "Brian Li"
date: "2019/9/1"
output: 
  html_document: 
    keep_md: yes
---

keep_md is set to yes to render a formatted markdown in github view.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calling libraries

```{r calling libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(randomNames)
library(dplyr)
library(stringr)
library(kableExtra)
```

We will circle around the usages of muate_at, str_detect and str_replace functions, and discuss how they can be used together to perform data transformation on string columns that contain some number data in some big data databases e.g. Hadoop. 

## Data Creation

The following code chunk will simulate the format and style of data that needs this treatment. It basically has a lot of string columns that contain number data with trailing zeros. However, some of them are meant to be purely alpha characters and remain as string columns. Personal ID (e.g. "P9455837B") and name (e.g. "Betty") are good examples of those.

```{r data creation}
# Set seed for repeatibility
set.seed(68614)

# Input the number of rows
row_num <- 30
# Input the number of columns
col_num <- 10

# Create base dataframe by randomly generating some names and ids
df <- data.frame(name = randomNames(row_num), 
                 id = paste0(sample(LETTERS,row_num, replace = TRUE), sample(c(1:1000000), row_num), sample(LETTERS,row_num, replace = TRUE)))

# Create abc_table
abc_table <- replicate(col_num, {
  
  # Number of rows with trailing zeros
  n<-sample(1:row_num,1)
  
  # Randomly generate the column
  col_vec <- c(str_pad(sample(1:10000,n, replace = TRUE), 9, side = "left", pad = "0"),
               sample(c("", NA, "N", "O", "P", "S"), row_num - n, replace = TRUE)
               )
  
  return(col_vec)
  
})

# Above replicate function will replcate the random column generating function for whatever col_num times (with each replication using a different seed)

# Convert matrix to dataframe
abc_table <- as.data.frame(abc_table)

# Assign colnames to abc_table
colnames(abc_table) <- paste0("attr_", c(1:ncol(abc_table)))

# Create def_table
def_table <- replicate(col_num, {
  
  # Number of rows with trailing zeros
  n<-sample(1:row_num,1)
  
  # Randomly generate the column
  col_vec <- c(str_pad(sample(1:10000,n, replace = TRUE), 9, side = "left", pad = "0"),
               sample(c("", NA, "N", "O", "P", "S"), row_num - n, replace = TRUE)
               )
  
  return(col_vec)
  
})

# Convert matrix to dataframe
def_table <- as.data.frame(def_table)

# Assign column names to def_table
colnames(def_table) <- paste0("ada_", c(1:ncol(def_table)))

df <- cbind(df,abc_table, def_table)

# Create 2 additional columns that have only character strings and alpha-numeric strings
df <- df %>% mutate(ix_1 = sample(c("East", "South", "West", "North"), row_num, replace = TRUE),
                    ix_2 = sample(c("1A2B", "3C4D", "5E6F", "8G9H"), row_num, replace = TRUE))

```

Hover the mouse over the table and scroll up-down and left-right to take a quick glimpse of the simulated data.

```{r kable}
knitr::kable(df) %>% kable_styling() %>% scroll_box(width = "100%", height = "600px")
```

## Core Process
First, create a function that removes leading zero strategically. That is if the string contains all 0 digits, I want it to return a single 0, else transform the 0s into empty string.

```{r remove zeros}

# Define a function that trims leading zeros
lead_zero_trim <- function(x){
  x_ws_trim <- trimws(as.character(x))
  ifelse(
    str_detect(x_ws_trim, "^0+$"), 
    str_replace(x_ws_trim, "^0+$", "0"), 
    str_replace(x_ws_trim,"^0+","")
    )
}

# Check if the function works
(test <- c("  0010  ", "   678910  ", "0000000", "    0000000", "0000100000000", "000000.9910"))
lead_zero_trim(test)

```

Then, mutate at columns that start with attr_ or ada_ by detecting matches using regular expression. 

```{r muate at columns}

# Uses tidyverse functions
test <- df %>% mutate_at(vars(matches("^(attr_|ada_|ix_)")),lead_zero_trim)

# See how df looks like now
knitr::kable(test) %>% kable_styling() %>% scroll_box(width = "100%", height = "600px")

```

You can see that strings that are ("", NA, or character only) remain intact as before. The function only transform those with leading zeros.<br>

Now, we would want to convert these digit strings into numeric columns.

```{r conversion to numeric, warning = FALSE}

# Dot . inherits the object after the pipe. i.e. individual column
# Wrap the result from lead_zer0_trim(.) with as.numeric and turn the process into a temp function by funs()
test <- df %>% mutate_at(vars(matches("^(attr_|ada_|ix_)")),funs(as.numeric(lead_zero_trim(.))))

# See how df looks like now
knitr::kable(test) %>% kable_styling() %>% scroll_box(width = "100%", height = "600px")

```

Oops! The as.numeric() turn those strings into NA, which we don't want.<br>
That means we have to transform those mixed-in chacters (e.g. N, O, P, S) into NA or some meaningful numbers. Then, depending on whether the column contains pure digits, apply as.numeric on them, but skip the ones containing only characters.

```{r check if data is a pure number, warning = FALSE}

# Define another function to check if it contains at least 1 row of proper number
is.proper.number <- function(x){
  str_detect(x, "^[0-9]*(\\.[0-9]*)?$")
}

# Check if it works
test <- c("1000", "0.1000", "10002", "123.444", "123.")
is.proper.number(test)

```
Putting all together.
```{r 2nd attempt to convert to numeric}

# Combine all these functions into 1 wrangle function
ig_wrangle <- function(x){
  x <- lead_zero_trim(x)
  
  # Lookups based on data dictionary
  x <- case_when(x == "N" ~ "1",
                 x == "O" ~ "99",
                 x == "P" ~ "999",
                 x == "S" ~ "9999",
                 # Turn those blank strings and true NA into NA_character_
                 # NA_character_ is used to enforce same type of data across entire column
                 x == "" | x == " " | is.na(x) ~ NA_character_,
                 # Else Case
                 TRUE ~ x)
  
  # Convert to numeric if there is at least one proper number data in that column
  if(any(is.proper.number(x))){
    as.numeric(x)
  } else {
    x
  }
  
  return(x)
}

test <- df %>% mutate_at(vars(matches("^(attr_|ada_|ix_)")),ig_wrangle)

# See how df looks like now
knitr::kable(test) %>% kable_styling() %>% scroll_box(width = "100%", height = "600px")


```
You can see that ix_1 and ix_2 remain as character string instead of turning into NA. The N, O, P, S are converted to meaningful numbers, with the missing still defined as missing. 

## Creating a holistic function
```{r function recap}

# Define a function that trims leading zeros
lead_zero_trim <- function(x){
  # Pass as.character in case it is a factor
  x_ws_trim <- trimws(as.character(x))
  ifelse(
    str_detect(x_ws_trim, "^0+$"), 
    str_replace(x_ws_trim, "^0+$", "0"), 
    str_replace(x_ws_trim,"^0+","")
    )
}

# Define another function to check if it contains at least 1 row of proper number
is.proper.number <- function(x){
  str_detect(x, "^[0-9]*(\\.[0-9]*)?$")
}

# Define ig_wrangle function
ig_wrangle <- function(x){
  x <- lead_zero_trim(x)
  
  # Lookups based on data dictionary
  x <- case_when(x == "N" ~ "1",
                 x == "O" ~ "99",
                 x == "P" ~ "999",
                 x == "S" ~ "9999",
                 # Turn those blank strings and true NA into NA_character_
                 # NA_character_ is used to enforce same type of data across entire column
                 x == "" | x == " " | is.na(x) ~ NA_character_,
                 # Else Case
                 TRUE ~ x)
  
  # Convert to numeric if there is at least one proper number data in that column
  if(any(is.proper.number(x))){
    as.numeric(x)
  } else {
    x
  }
  
  return(x)
}

# Create a wrapper to pass dataframe and column selection based on regex starts with
ig_df_wrangle <- function(df, start_with_str){

  df <- df %>% mutate_at(vars(matches(paste0("^(",start_with_str,")"))),ig_wrangle)
  
}

# Test
test2 <- ig_df_wrangle(df, "attr_|ada_|ix_")
# This will apply the ig_wrangle function to columns that starts with attr_, ada_ or ix_

```

